SUPPORTING INFORMATION:\\ Headgroup Structure and Cation Binding in Phosphatidylserine Lipid Bilayers

### 2019-09-04 Manuscript/manuscriptPS.tex
Structure without additional ions

  Additional simulation details are given in the supplementary information.
  $^*$ Force field parameters for PS lipids generated for this work, for full details see the supplementary information.

@@

  The additional simulation details are given in the supplementary information.

@@

POPC:POPS (5:1)/Na$^{+}$  & GROMOS-CKP\footnote{Force field parameters for PS lipids generated for this work, for full details see the supplementary information.} \cite{piggot12}  & 0  & 110:22 & 4620  & 0  & 298  & 2$\times$500  & 2$\times$100  & \citenum{POPCpopsGROMOSCKPwithNa} \tabularnewline

@@

\subsection{Headgroup and glycerol backbone C--H bond order parameters of POPS from $^{13}$C NMR}
The INEPT and 2D R-PDLF experiments from POPS samples gave well resolved spectra for all the
carbons in the headgroup and glycerol backbone regions (Fig. \ref{PShgSIGNSsimpson}).
The glycerol backbone carbon peaks were assigned according to the POPC spectra~\cite{ferreira13}, whereas
the peaks for $\beta$ and $\alpha$ carbons were assigned according to the
known C--H bond order parameters from the $^2$H\,NMR experiments~\cite{browning80}.
Slices of the R-PDLF spectra and the resulting  $S_{\rm CH}$ values
are shown in the supplementary information (Fig. \ref{DPslices}).

@@

The different PS MD models produce highly varied headgroup and glycerol backbone  $S_{\rm CH}$ (Fig.~\ref{HGorderParametersPS})
and structures (Fig.~\ref{HGandGLYstructuresPS} and section~\ref{Diheds} in the supplementary information).
As was previously observed for PC lipids \cite{botan15}, also none of the PS models produces a set of  $S_{\rm CH}$ in full quantitative agreement with the experiments.
In fact, the models perform less well for PS than for PC
(Figs.~\ref{HGorderParametersPS} and \ref{comparisonTablePS} vs. Figs.~2 and 4 in Ref.~\cite{botan15}),
which complicates the interpretation of structural differences between the PC and PS. However, concentrating on the headgroup alone, we see that the best performing models (Slipids, CHARMM36 and CHARMM36-UA) indeed do replicate the larger-than-in-PC
forking of the $\alpha$-carbon that is observed in experiments (Fig.~\ref{HGorderParametersPS}).
Additionally, the Slipids force field correctly produces the significantly smaller $\beta$-carbon  $S_{\rm CH}$
for PS compared to PC (Fig. \ref{HGorderParametersPS} vs. Fig. 2 in Ref.~\citenum{botan15}), in line with
experiments (Fig. \ref{HGorderParameters}).

\begin{figure}[]
  \centering
  \includegraphics[width=9.0cm]{../Figs/HGorderparametersPS.pdf}
  \caption{\label{HGorderParametersPS}
    C--H bond order parameters, $S_{\rm CH}$, of PS headgroup ($\beta$ and $\alpha$) and
    glycerol backbone (g$_3$, g$_2$, g$_1$) carbons from NMR experiments (horizontal lines),
    and MD simulations with different force fields (symbols).
    Experimental data for DOPS are measured with 0.1~M of NaCl~\cite{browning80},
    while all the other data are with counterions only.
    The data for DOPS is at 303~K and for POPS at 298~K.
    Light blue areas span 0.04 units around the average of the extremal experimental values,
    in accordance with the expected quantitative accuracy of experiments~\cite{ollila16}.
    The vertical bars shown for all simulation values (excl. MacRog~K$^+$)
    are not error bars, but demonstrate that for these systems
    we had at least two data sets; the ends of the bars mark the extreme values
    from the sets, and the symbol marks their measurement-time-weighted average.
  }
\end{figure}

\begin{figure}[]
  \centering
  \includegraphics[width=9.0cm]{../Figs/structures.eps}
  \caption{\label{HGstructuresPSandPC}
    Overlayed snapshots from simulations conducted with CHARMM36---the force field producing the best agreement with experiments---demonstrate the conformational fluctuations
    of lipid headgroups.
    (A) Overlaying  C$_\gamma$-C$_\beta$-C$_\alpha$ carbons demonstrate fluctuations around O$_\gamma$-C$_\gamma$-C$_\beta$-C$_\alpha$, C$_\gamma$-C$_\beta$-C$_\alpha$-O$_\alpha$,
    and N-C$_\beta$-C$_\alpha$-O$_\alpha$ dihedrals of the PS headgroup and
    (B) overlaying C$_\beta$-C$_\alpha$-O$_\alpha$ carbonds demonstrate fluctuations around N-C$_\beta$-C$_\alpha$-O$_\alpha$ and C$_\beta$-C$_\alpha$-O$_\alpha$-P
    dihedrals of the PS and PC headgroups. The trajectory used for CHARMM36 POPC is available at Ref.~\citenum{POPCcharmm36T303K}.}
\end{figure}

\begin{figure}[]
  \centering
  \includegraphics[width=9.0cm]{../Figs/comparisonTablePS.pdf}
  \caption{\label{comparisonTablePS}
    Rough subjective ranking of force fields based on how far are the simulated order parameters from the experimental
    values ({\textsf{\small M}} indicates a magnitude problem) and how well the difference between order parameters
    of the two hydrogens attached in the same carbon are reproduced ({\textsf{\small F}} indicates a forking problem)
    in Figure~\ref{HGorderParametersPS}.
    The letter size increases with problem severity. Color scheme: "within experimental error" (dark green), "almost within experimental error" (light green), "clear deviation from experiments" (light red), and "major deviation from experiments" (dark red). The $\Sigma$-column shows the total deviation of the force field, when individual carbons are given weights of 0 (matches experiment), 1, 2, and 4 (major deviation). For full details of the assessment, see Supplementary Information.
  }
\end{figure}

Interestingly, in the three models that best fit the experimental data,
the C$_\alpha$-C$_\beta$-C$_\gamma$-O$_\gamma$ dihedral angle distribution has a single peak around 120$^{\circ}$, while
other models yield binodal distributions (Fig. \ref{dihedralsHG}).
The restricted motion is also visible in the sampled conformations (Figs.~\ref{HGstructuresPSandPC} (A) and \ref{HGandGLYstructuresPS}),
suggesting that the rotation of the carboxyl group is limited in the serine headgroup.
On the other hand, the C$_\beta$-C$_\alpha$-O$_\alpha$-P dihedral angle rotates relatively
freely between approximately 100$^{\circ}$ and 300$^{\circ}$ in the best three models (as seen also in Fig.~\ref{HGstructuresPSandPC} (B)),
while other models yield more restricted conformations (Fig. \ref{dihedralsHG}).

In simulations that have the best agreement with experiments (CHARMM36 and Slipids),
the N-C$_\beta$-C$_\alpha$-O$_\alpha$ dihedral exhibits a more asymmetric
angle distribution for PS than for PC headgroup (Figs.~\ref{HGstructuresPSandPC} (B) and \ref{dihedralsHGpc}),
which might reflect the increased rigidity proposed in the early experimental studies \cite{browning80,buldt81}.
The suggested characteristic conformations of the PS headgroup can be useful when interpreting experiments,
however, as the none of the tested models fully reproduces the experimental order parameters, more accurate MD force fields are required to elucidate the correct conformational ensemble.

>>>>

  Additional simulation details are given in the Supporting Information.
  $^*$ Force field parameters for PS lipids generated for this work, for full details see the Supporting Information.

@@

  The additional simulation details are given in the Supporting Information.

@@

POPC:POPS (5:1)/Na$^{+}$  & GROMOS-CKP\footnote{Force field parameters for PS lipids generated for this work, for full details see the Supporting Information.} \cite{piggot12}  & 0  & 110:22 & 4620  & 0  & 298  & 2$\times$500  & 2$\times$100  & \citenum{POPCpopsGROMOSCKPwithNa} \tabularnewline

@@

\subsection{Headgroup and glycerol backbone C--H bond order parameters of POPS from $^{13}$C NMR}
The INEPT and 2D R-PDLF experiments from POPS samples gave well resolved spectra for all the
carbons in the headgroup and glycerol backbone regions (Fig. \ref{PShgSIGNSsimpson}).
The glycerol backbone carbon peaks were assigned according to the POPC spectra~\cite{ferreira13}, whereas
the peaks for $\beta$ and $\alpha$ carbons were assigned according to the
known C--H bond order parameters from the $^2$H\,NMR experiments~\cite{browning80}.
Slices of the R-PDLF spectra and the resulting  $S_{\rm CH}$ values
are shown in the Supporting Information (Fig. \ref{DPslices}).

@@

The different PS MD models produced highly varied headgroup and glycerol backbone  $S_{\rm CH}$ (Fig.~\ref{HGorderParametersPS})
and structures (Fig.~\ref{HGandGLYstructuresPS} and section~\ref{Diheds} in the Supporting Information).
As was previously observed for PC lipids \cite{botan15}, also none of the PS models produced a set of  $S_{\rm CH}$ in full quantitative agreement with the experiments.
In fact, the models perform less well for PS than for PC
(Figs.~\ref{HGorderParametersPS} and \ref{comparisonTablePS} vs. Figs.~2 and 4 in Ref.~\citenum{botan15}),
which complicates the interpretation of structural differences between the PC and PS. However, concentrating on the headgroup alone, we see that the best performing models (Slipids, CHARMM36, and CHARMM36-UA) indeed do qualitatively replicate the larger-than-in-PC
forking of the $\alpha$-carbon (these models give forkings of $\sim\!0.01$ or smaller for PC~\cite{botan15} and between $\sim\!0.05$ and $\sim\!0.12$ for PS) that is observed in experiments
(Fig.~\ref{HGorderParameters}).
Additionally, the Slipids force field correctly produces the significantly smaller $\beta$-carbon order parameter
for PS ($S^\beta_{\rm CH}\approx-0.09$, see Fig.~\ref{HGorderParametersPS}) compared to PC ($S^\beta_{\rm CH}\approx-0.03$,~see Fig.~2 in Ref.~\citenum{botan15}),
%Fig. \ref{HGorderParametersPS} vs. Fig. 2 in Ref.~\citenum{botan15})
in line with experiments (Fig. \ref{HGorderParameters}).

\begin{figure}[]
  \centering
  \includegraphics[width=9.0cm]{../Figs/HGorderparametersPS.pdf}
  \caption{\label{HGorderParametersPS}
    C--H bond order parameters, $S_{\rm CH}$, of the PS headgroup ($\beta$ and $\alpha$) and
    glycerol backbone (g$_3$, g$_2$, g$_1$) carbons from NMR experiments (horizontal lines),
    and MD simulations with different force fields (symbols).
    Experimental data for DOPS had been measured with 0.1\,M of NaCl~\cite{browning80},
    while all the other data are with counterions only.
    The data for DOPS is at 303~K and for POPS at 298~K.
    The light blue areas span 0.04 units around the average of the extremal experimental values,
    in accordance with the expected quantitative accuracy of experiments~\cite{ollila16}.
    The vertical bars shown for all simulation values (excl. MacRog~K$^+$)
    are not error bars, but demonstrate that for these systems
    we had at least two data sets; the ends of the bars mark the extreme values
    from the sets, and the symbol marks their measurement-time-weighted average.
  }
\end{figure}

\begin{figure}[]
  \centering
  \includegraphics[width=9.0cm]{../Figs/comparisonTablePS.pdf}
  \caption{\label{comparisonTablePS}
    Rough subjective ranking of force fields based on how far the simulated order parameters were from the experimental
    values ({\textsf{\small M}} indicates a magnitude problem) and how well the difference between order parameters
    of two hydrogens attached to the same carbon were reproduced ({\textsf{\small F}} indicates a forking problem)
    in Fig.~\ref{HGorderParametersPS}.
    The letter size increases with problem severity. Color scheme: "within experimental error" (dark green), "almost within experimental error" (light green), "clear deviation from experiments" (light red), and "major deviation from experiments" (dark red). The $\Sigma$-column shows the total deviation of the force field, when individual carbons are given weights of 0 (matches experiment), 1, 2, and 4 (major deviation). For full details of the assessment, see section~\ref{Ranking} in the Supporting Information.
  }
\end{figure}

\begin{figure}[]
  \centering
  \includegraphics[width=9.0cm]{../Figs/structures.eps}
  \caption{\label{HGstructuresPSandPC}
    Overlayed snapshots from simulations conducted with CHARMM36---the force field producing the best agreement with experiments---demonstrate the conformational fluctuations
    in lipid headgroups.
    (A) Overlaying the C$_\gamma$, C$_\beta$, and C$_\alpha$ carbons demonstrates fluctuations around the O$_\gamma$-C$_\gamma$-C$_\beta$-C$_\alpha$, C$_\gamma$-C$_\beta$-C$_\alpha$-O$_\alpha$,
    and N-C$_\beta$-C$_\alpha$-O$_\alpha$ dihedrals of the PS headgroup.
    (B) Overlaying the C$_\beta$, C$_\alpha$, and O$_\alpha$ atoms demonstrates fluctuations around the N-C$_\beta$-C$_\alpha$-O$_\alpha$ and C$_\beta$-C$_\alpha$-O$_\alpha$-P
    dihedrals of the PS and PC headgroups. The trajectory used for CHARMM36 POPC is available at Ref.~\citenum{POPCcharmm36T303K}.}
\end{figure}

Interestingly, in the three models that best fit the experimental data,
the C$_\alpha$-C$_\beta$-C$_\gamma$-O$_\gamma$ dihedral angle distribution has a single peak around 120$^{\circ}$, while
the other models yield binodal distributions (Fig. \ref{dihedralsHG}).
The restricted motion is also visible in the sampled conformations (Figs.~\ref{HGstructuresPSandPC} (A) and \ref{HGandGLYstructuresPS}),
suggesting that the rotation of the carboxyl group is limited in the serine headgroup.
On the other hand, the C$_\beta$-C$_\alpha$-O$_\alpha$-P dihedral angle rotates relatively
freely between approximately 100$^{\circ}$ and 300$^{\circ}$ in the best three models (as seen also in Fig.~\ref{HGstructuresPSandPC} (B)),
while other models yield more restricted conformations (Fig. \ref{dihedralsHG}).

In simulations that have the best agreement with experiments (CHARMM36 and Slipids),
the N-C$_\beta$-C$_\alpha$-O$_\alpha$ dihedral exhibits a more asymmetric
angle distribution for PS than for PC headgroup (Figs.~\ref{HGstructuresPSandPC} (B) and \ref{dihedralsHGpc}),
which might reflect the increased rigidity proposed in the early experimental studies \cite{browning80,buldt81}.
Indeed, the characteristic conformations of the PS headgroup suggested here can be useful when interpreting experiments---however,
as none of the tested models fully reproduces the experimental order parameters, more accurate MD force fields are required to reveal the true conformational ensemble.

### 2019-09-04 Manuscript/manuscriptPSsuppl.tex
Structure without additional ions

\title{ Supporting Information:\\ Headgroup structure and cation binding in phosphatidylserine lipid bilayers}

@@

\pagebreak
\section{Dihedral angle distributions of the headgroup and glycerol backbone
  regions of PS lipids from different simulation models}\label{Diheds}

The dihedral angles and structures of the glycerol backbone and headgroup regions of POPS lipids show
wide variety between different simulation models (Figs.~\ref{dihedralsHG}, \ref{dihedralsGLY} and \ref{HGandGLYstructuresPS}).
Detailed discussion of the structural differences is limited by the lack of realistic model that would correctly reproduce the
headgroup and glycerol backbone order parameters. However, some structural characteristics
of PS headgroup can be suggested based on the best available models
(Figs.~\ref{dihedralsHGpc}, \ref{dihedralsGLYpc} and \ref{HGandGLYstructuresPSPC}), as discussed in the main text.
The glycerol backbone structures are not discussed in this work because our focus is on PS headgroup.
However, the data presented here can be useful for future investigations.
%backbone order parameters of C$_2$ and C$_3$ from Slipids simulations
%differ significantly from the other simulation results and experiments (Fig. \ref{HGorderParametersPS}),
%as observed previously also for PC lipids \cite{botan15}.
%The origin of this difference is more difficult to track without more elaborate analysis,
%because different models show very complicated patterns of distinct structures
%in the glycerol backbone region (Figs.~\ref{HGstructuresPS} and \ref{dihedralsHG}).

\begin{figure*}[]
  \centering
  \includegraphics[width=8.0cm]{../Figs/PS_Labels.png}
  \includegraphics[width=16.0cm]{../Figs/figS7.png}
  \caption{\label{dihedralsHG}
    Dihedral angle distributions of the headgroup region of POPS from different simulation models.
  }
\end{figure*}

\begin{figure*}[]
  \centering
    \includegraphics[width=8.0cm]{../Figs/PS_Labels.png}
  \includegraphics[width=13.0cm]{../Figs/figS6.png}
  \caption{\label{dihedralsGLY}
    Dihedral angle distributions of the glycerol backbone region of POPS lipids from different simulation models.
  }
\end{figure*}

\begin{figure*}[]
  \centering
  \includegraphics[width=18.0cm]{../Figs/figS8.png}
  \caption{\label{HGandGLYstructuresPS}
    Overlayed snapshots of the headgroup (C$_\gamma$-C$_\beta$-C$_\alpha$ carbons overlayed) and
    glycerol backbone (C$_{\rm g1}$-C$_{\rm g2}$-C$_{\rm g3}$ carbons overlayed) regions from different POPS simulations.
    Note that only one of the two O$_\gamma$ atoms in the carboxylate group from PS lipids is shown.
  }
\end{figure*}

\begin{figure*}[]
  \centering
  \includegraphics[width=16.0cm]{../Figs/figS11.png}
  \caption{\label{dihedralsHGpc}
    Dihedral angle distributions of the headgroup regions from CHARMM36 and Slipids simulations
    compared between the POPC and POPS lipids. 
    The CHARMM36 POPC simulation is from Ref.~\citenum{POPCcharmm36T303K} and Slipids POPC from Ref.~\citenum{POPCslipids298K}.
  }
\end{figure*}

\begin{figure*}[]
  \centering
  \includegraphics[width=15.0cm]{../Figs/figS10.png}
  \caption{\label{dihedralsGLYpc}
    Dihedral angle distributions of the glycerol backbone regions from CHARMM36 and Slipids simulations
    compared between the POPC and POPS lipids.
    The CHARMM36 POPC simulation is from Ref.~\citenum{POPCcharmm36T303K} and Slipids POPC from Ref.~\citenum{POPCslipids298K}.
  }
\end{figure*}

\begin{figure}[]
  \centering
  \includegraphics[width=9.0cm]{../Figs/figS8_POPC.png}
  \caption{\label{HGandGLYstructuresPSPC}
    Overlayed snapshots of the headgroup and glycerol backbone regions 
    from CHARMM36 and Slipids simulations compared between the POPC and POPS lipids.
    The CHARMM36 POPC simulation is from Ref.~\citenum{POPCcharmm36T303K} and Slipids POPC from Ref.~\citenum{POPCslipids298K}.
  }
\end{figure}

>>>>

\title{SUPPORTING INFORMATION:\\ Headgroup Structure and Cation Binding in Phosphatidylserine Lipid Bilayers}

@@

\pagebreak
\section{Dihedral angle distributions in headgroup and glycerol backbone
  regions of PS lipids from different simulation models}\label{Diheds}

The dihedral angle distributions (Figs.~\ref{dihedralsHG} and~\ref{dihedralsGLY}) and structures (Fig.~\ref{HGandGLYstructuresPS}) of the glycerol backbone and headgroup regions of POPS lipids showed a
wide variety between different MD force fields.
Detailed discussion on the relevance of the observed structural differences is limited by the lack of a realistic force field (that would correctly reproduce the
headgroup and glycerol backbone C--H bond order parameters, $S_\mathrm{CH}$). However, some structural characteristics
of the PS headgroup can be suggested based on the best available models for it
(Figs.~\ref{dihedralsHGpc}, \ref{dihedralsGLYpc} and \ref{HGandGLYstructuresPSPC}), as discussed in the main text.
The glycerol backbone structures are not discussed in this work, because our focus is on the PS headgroup.
However, the data presented here can be useful for future investigations.
%backbone order parameters of C$_2$ and C$_3$ from Slipids simulations
%differ significantly from the other simulation results and experiments (Fig. \ref{HGorderParametersPS}),
%as observed previously also for PC lipids \cite{botan15}.
%The origin of this difference is more difficult to track without more elaborate analysis,
%because different models show very complicated patterns of distinct structures
%in the glycerol backbone region (Figs.~\ref{HGstructuresPS} and \ref{dihedralsHG}).

\begin{figure*}[]
  \centering
  \includegraphics[width=0.5\textwidth]{../Figs/PS_Labels.png}
  \includegraphics[width=0.8\textwidth]{../Figs/figS7.png}
  \caption{\label{dihedralsHG}
    Dihedral angle distributions in the headgroup region of POPS from different simulation models.
  }
\end{figure*}

\begin{figure*}[]
  \centering
    \includegraphics[width=0.5\textwidth]{../Figs/PS_Labels.png}
  \includegraphics[width=0.8\textwidth]{../Figs/figS6.png}
  \caption{\label{dihedralsGLY}
    Dihedral angle distributions in the glycerol backbone region of POPS from different simulation models.
  }
\end{figure*}

\begin{figure*}[]
  \centering
  \includegraphics[width=1.03\textwidth]{../Figs/figS8.png}
  \caption{\label{HGandGLYstructuresPS}
    Overlayed snapshots of the headgroup (C$_\gamma$-C$_\beta$-C$_\alpha$ carbons overlayed) and
    glycerol backbone (C$_{\rm g_1}$-C$_{\rm g_2}$-C$_{\rm g_3}$ carbons overlayed) regions from POPS simulations with different force fields.
    Note that only one of the two O$_\gamma$ atoms in the carboxylate group of PS lipids is shown.
  }
\end{figure*}

\begin{figure*}[]
  \centering
  \includegraphics[width=0.9\textwidth]{../Figs/figS11.png}
  \caption{\label{dihedralsHGpc}
    Dihedral angle distributions in the headgroup region from CHARMM36 and Slipids simulations
    compared between the POPC and POPS lipids. 
    The CHARMM36 POPC simulation is from Ref.~\citenum{POPCcharmm36T303K} and Slipids POPC from Ref.~\citenum{POPCslipids298K}.
  }
\end{figure*}

\begin{figure*}[]
  \centering
  \includegraphics[width=0.9\textwidth]{../Figs/figS10.png}
  \caption{\label{dihedralsGLYpc}
    Dihedral angle distributions in the glycerol backbone region from CHARMM36 and Slipids simulations
    compared between the POPC and POPS lipids.
    The CHARMM36 POPC simulation is from Ref.~\citenum{POPCcharmm36T303K} and Slipids POPC from Ref.~\citenum{POPCslipids298K}.
  }
\end{figure*}

\begin{figure}[]
  \centering
  \includegraphics[width=0.9\textwidth]{../Figs/figS8_POPC.png}
  \caption{\label{HGandGLYstructuresPSPC}
    Overlayed snapshots of the headgroup and glycerol backbone regions 
    from CHARMM36 and Slipids simulations compared between the POPC and POPS lipids.
    The CHARMM36 POPC simulation is from Ref.~\citenum{POPCcharmm36T303K} and Slipids POPC from Ref.~\citenum{POPCslipids298K}.
  }
\end{figure}
